Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation

### 2020-08-26 summary.Rmd
still working on additional example

```{r}
import::from(magrittr, `%>%`)
library(plot.matrix)
library(ggplot2)

@@

*Proof*
Let $\Lambda^{(k)} = 
\begin{bmatrix} \lambda^{(k,1)} & \cdots & \lambda^{(k, K)} \end{bmatrix}
\in \mathbb{R}^{n_k \times K}$.  
Let $X$ be a block diagonal matrix 
$X = \diag(\Lambda^{(1)}, ..., \Lambda^{(K)}) \in \mathbb{R}^{n \times K^2}$.  
Let $L^{(k)}$ be a block diagonal matrix of column vectors $\lambda^{(lk)}$ for 
$l = 1, ..., K$. $L^{(k)} = \diag(\lambda^{(1k)}, ..., \lambda^{(Kk)}) \in 
\mathbb{R}^{n \times K}$.  
Let $Y = \begin{bmatrix} L^{(1)} & \cdots & L^{(K)} \end{bmatrix} \in 
\mathbb{R}^{n \times K^2}$.  
Then $P = X Y^\top$.  
Similar to the $K = 2$ case, we again have $Y = X \Pi$ for a permutation matrix
$\Pi$, so $P = X \Pi X^\top$.  
The permutation described by $\Pi$ has $K$ fixed points, which correspond to 
$K$ eigenvalues equal to $1$ with corresponding eigenvectors $e_k$ where 
$k = r (K + 1) + 1$ for $r = 0, ..., K - 1$. It also has 
$\binom{K}{2} = K (K - 1) / 2$ cycles of order $2$. Each cycle corresponds to 
a pair of eigenvalues $+1$ and $-1$ and a pair of eigenvectors 
$(e_s + e_t) / \sqrt{2}$ and $(e_s - e_t) / \sqrt{2}$.

@@

**Theorem 5**. Under assumptions 1-6 as laid out by \citet{307cbeb9b1be48299388437423d94bf1}, the root mean square errors of the 
estimators outputted by algorithm 4 converge to the true values as follows:

$$\sqrt{\frac{1}{n K} \sum_{k, l = 1}^K 
||\hat{\lambda}^{(kl)} - \lambda^{(kl)}||^2} 
\stackrel{P}{\to} 0$$

*Proof*. ***TODO***

The convergence rate of algorithm 4 is the same as that of the methods 
described by \citet{307cbeb9b1be48299388437423d94bf1}.

\begin{comment}

### Using Adjacency Spectral Embedding

For this section, we will focus on the case $K = 2$. Under this condition, the 
PABM is equivalent to the GRDPG with signature $(3, 1)$.

The adjacency spectral embeddings of both $A$ and $P$ are not unique. In 
particular, let $Z$ be the ASE of $P$. Then $P = Z I_{3, 1} Z^\top$. However, 
for any $Q \in \mathbb{O}(3, 1)$, 
$(ZQ) I_{3, 1} (ZQ)^\top = Z (Q I_{3, 1} Q^\top) Z^\top = Z I_{3, 1} Z^\top = P$,
so $ZQ$ is also a valid ASE Of $P$. If we can find $Q \in \mathbb{O}(3, 1)$ 
such that $ZQ = XU$, we can compute the parameters $\{\lambda^{(kl)}\}$ 
directly. Furthermore, if we instead use the ASE of the adjacency matrix $A$, 
$\exists Q \in \mathbb{O}(3, 1)$ such that $\max\limits_i ||Q \hat{z}_i - XU||$ 
is minimized (and goes to zero under a probabilistic model for
$\{\lambda^{(kl)}\}$'s).

Thus if we can identify $Q$ such that the $ZQ = XU$ in the case of embedding 
$P$ or $\hat{Z} Q - XU$ is minimized in the case of embedding $A$, we can 
identify or estimate the PABM parameters directly from the embedding. $XU$ is 
unknown, but we can still obtain an embedding $ZQ$ that follows the properties 
of $XU$, which will yield estimates $\{\hat{\lambda}^{(kl)}\}$ that are valid 
in that they will produce the edge probability matrix $P$.[^lambdas]

[^lambdas]: Note that the set $\{\lambda^{(kl)}\}_K$ that produces a unique 
$P$ is not unique \citep{noroozi2019estimation}.

In particular, we can note the following properties of $XU$:

1. If row $i$ is in cluster 1, then its second index is 0. If it is in cluster 
2, its first index is 0.
2. If row $i$ is in cluster 1, then its third and fourth indices are equal. If 
it is in cluster 2, the fourth index is the negative of its third index.
3. If rows $i$ and $j$ are in different clusters, their dot product is 0.

If we have cluster memberships (either known *a priori* or estimated using a 
clustering method), then we can estimate $XU$ by picking
$Q \in \mathbb{O}(3, 1)$ such that $ZQ$ best fits these properties.

$\mathbb{O}(3, 1)$ happens to be the Lorentz group, and each 
$Q \in \mathbb{O}(3, 1)$ can be represented as the product of six matrices that 
depend on a total of four parameters \cite{10.1088/978-1-6817-4254-0}. This 
lends itself as an optimization problem:

\begin{equation}
\begin{split}
  \min_{\theta, \phi, \psi, \tau} & ||\xi^{(1, 2)}||^2 + ||\xi^{(2, 1)}||^2 + 
  ||\xi^{(1, 3)} - \xi^{(1, 4)}||^2 + ||\xi^{(2, 3)} + \xi^{(2, 4)}||^2 \\
  \text{s.t. } & \Xi = Z Q(\theta, \phi, \psi, \tau) \\
  & \xi^{(i, j)} \text{ is the } j^{th} \text{ column of } \Xi \text{ with rows from the } i^{th} \text{ cluster}
\end{split}
\end{equation}

Note that if $Q$ can be properly estimated, the asymptotic results from 
\citet{rubindelanchy2017statistical} can be applied here.

\end{comment}

@@

# Real data example

@@

Comparing against the ground truth species labels, Algorithm 2 achieves an 
accuracy of 63\% and an adjusted Rand index of `r ari`\%. In comparison, 
\citet{noroozi2019estimation} achieved an adjusted Rand index of 73\% using 
sparse subspace clustering.

```{r butterfly, fig.height = 4, fig.width = 4, fig.cap = 'Adjacency matrix of the Leeds Butterfly dataset after sorting by the clustering outputted by Algorithm 2.', cache = TRUE}
library(plot.matrix)
plot(A.star, border = NA, key = NULL, 
     xlab = '', ylab = '', main = '',
     axis.row = NULL, axis.col = NULL, asp = 1)
```

>>>>

```{r}
import::from(magrittr, `%>%`)
library(plot.matrix)
library(ggplot2)
library(plot.matrix)

@@

*Proof*. 
Let $\Lambda^{(k)} = 
\begin{bmatrix} \lambda^{(k,1)} & \cdots & \lambda^{(k, K)} \end{bmatrix}
\in \mathbb{R}^{n_k \times K}$.  
Let $X$ be a block diagonal matrix 
$X = \diag(\Lambda^{(1)}, ..., \Lambda^{(K)}) \in \mathbb{R}^{n \times K^2}$.  
Let $L^{(k)}$ be a block diagonal matrix of column vectors $\lambda^{(lk)}$ for 
$l = 1, ..., K$. $L^{(k)} = \diag(\lambda^{(1k)}, ..., \lambda^{(Kk)}) \in 
\mathbb{R}^{n \times K}$.  
Let $Y = \begin{bmatrix} L^{(1)} & \cdots & L^{(K)} \end{bmatrix} \in 
\mathbb{R}^{n \times K^2}$.  
Then $P = X Y^\top$.  
Similar to the $K = 2$ case, we again have $Y = X \Pi$ for a permutation matrix
$\Pi$, so $P = X \Pi X^\top$.  
The permutation described by $\Pi$ has $K$ fixed points, which correspond to 
$K$ eigenvalues equal to $1$ with corresponding eigenvectors $e_k$ where 
$k = r (K + 1) + 1$ for $r = 0, ..., K - 1$. It also has 
$\binom{K}{2} = K (K - 1) / 2$ cycles of order $2$. Each cycle corresponds to 
a pair of eigenvalues $+1$ and $-1$ and a pair of eigenvectors 
$(e_s + e_t) / \sqrt{2}$ and $(e_s - e_t) / \sqrt{2}$.

@@

**Theorem 5**. Under regularity and sparsity assumptions, and under the further 
assumption that $K$ is fixed,

$$\max_{k, l \in \{1, ..., K\}} 
||\hat{\lambda}^{(kl)} - \lambda^{(kl)}|| = 
O_P \bigg(\frac{(\log n_k)^c}{\sqrt{n_k}} \bigg)$$

*Proof (sketch)*.  
Let $P$ and $A$ be organized by cluster such that the elements of blocks 
$P^{(kl)}$ and $A^{(kl)}$ correspond to edges between clusters $k$ and $l$. 
By definition, $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. This implies 
that $P^{(kl)} = \sigma_{kl}^2 u^{(kl)} (v^{(kl)})^\top$ has one singular value 
and one pair of singular vectors. By convention, we can let 
$\lambda^{(kl)} = \sigma_{kl} u^{(kl)}$. Let 
$\hat{U}^{(kl)} \hat{\Sigma}^{(kl)} (\hat{V}^{kl})^\top$ be the singular value 
decomposition of $A^{(kl)}$, and let 
$\hat{\sigma}_{kl} \hat{u}^{(kl)} (\hat{v}^{(kl)})^\top$ be its one-dimensional 
approximation. Then 
$\hat{\lambda}^{(kl)} = \hat{\sigma}_{kl} \hat{u}^{(kl)}$.[^alg4] Thus 
$\hat{\lambda}^{(kl)}$ is a one-dimensional adjacency embedding approximation 
of $\lambda^{(kl)}$.  
Then by Theorem 5 from \citet{rubindelanchy2017statistical}, the adjacency 
embedding $\hat{\lambda}^{(kl)}$ approximates $\lambda^{(kl)}$ at rate 
$\frac{(\log n_k)^c}{n_k^{1/2}}$.

[^alg4]: Note that this estimator is similar to but slightly worse than the 
estimator from Algorithm 4.

@@

# Real data examples

@@

Comparing against the ground truth species labels, Algorithm 2 achieves an 
accuracy of 63\% and an adjusted Rand index of `r ari`\%. In comparison, 
\citet{noroozi2019estimation} achieved an adjusted Rand index of 73\% using 
sparse subspace clustering on the same dataset.

```{r}
plot.A <- function(A, margins = c(0, 0, 0, 0)) {
  default.mar <- par()$mar
  par(mar = margins)
  plot(A, border = NA, key = NULL, 
       xlab = '', ylab = '', main = '',
       axis.row = NULL, axis.col = NULL, asp = 1)
  par(mar = default.mar)
}
```

```{r butterfly, fig.height = 2, fig.width = 2, fig.cap = 'Adjacency matrix of the Leeds Butterfly dataset after sorting by the clustering outputted by Algorithm 2.', cache = TRUE}
plot.A(A.star)
```
